---
title: "steelhead_PCA_DAPC"
author: "Kimberly Ledger"
date: "4/12/2022"
output: github_document
---

# looking at O.mykiss population structure  

load libraries
```{r, message = FALSE}
library(adegenet) #this package is used for analysis of genetic/genomic data 
library(dplyr) # data manipulation
library(tidyr) # data manipulation
library(forcats)
library(ggplot2)
```


# Part 1: create genind objects

```{r}
#onmy_n_df <- read.csv("outputs/onmy_loci4genstr_outflank_hwe.csv")
onmy_n_df <- read.csv("outputs/omy_loci4genstr_outflank_hwe_oneadapt.csv")
```

here i will subset PRE and POST dam individuals 
```{r}
geno_pre <- onmy_n_df %>%
  filter(Time == "Pre") %>%
  dplyr::mutate(Location = fct_relevel(Location, "BD", "SBLR", "ID", "AD")) %>% 
  arrange(Location, rkm)

geno_post <- onmy_n_df %>%
  filter(Time == "Post") %>%
  dplyr::mutate(Location = fct_relevel(Location, "BD", "SBLR", "ID", "AD")) %>% 
  arrange(Location, rkm)
```

## create genind objects from pre and post dam samples
```{r}
#geno_n_pre <- geno_pre[,c(2:297)]        #for "outputs/onmy_loci4genstr_outflank_hwe.csv" use 297 
geno_n_pre <- geno_pre[,c(2:265)]        #for "outputs/omy_loci4genstr_outflank_hwe_oneadapt.csv" use 265

rownames(geno_n_pre) <- geno_pre$Sample_ID
col_geno_n_pre <- gsub("\\.", "_", colnames(geno_n_pre))
colnames(geno_n_pre) <- col_geno_n_pre

#meta_pre <- geno_pre[,-c(1:297)]
meta_pre <- geno_pre[,-c(1:265)]

pre_loci <- colnames(geno_n_pre)
pre_ind <- rownames(geno_n_pre)
pre_location <- meta_pre$Location
pre_rkm <- meta_pre$rkm
pre_site <- meta_pre$Sampling_Site

onmy_genind_pre_location <- df2genind(geno_n_pre,
                         sep="/",
                         ind.names=pre_ind,
                         loc.names=pre_loci, 
                         pop = pre_location,     
                         ploidy = 2)
onmy_genind_pre_site <- df2genind(geno_n_pre,
                         sep="/",
                         ind.names=pre_ind,
                         loc.names=pre_loci, 
                         pop = pre_site,      
                         ploidy = 2)
onmy_genind_pre_location
onmy_genind_pre_site


#geno_n_post <- geno_post[,c(2:297)]
geno_n_post <- geno_post[,c(2:265)]

rownames(geno_n_post) <- geno_post$Sample_ID
col_geno_n_post <- gsub("\\.", "_", colnames(geno_n_post))
colnames(geno_n_post) <- col_geno_n_post

#meta_post <- geno_post[,-c(1:297)]
meta_post <- geno_post[,-c(1:265)]

post_loci <- colnames(geno_n_post)
post_ind <- rownames(geno_n_post)
post_location <- meta_post$Location
post_rkm <- meta_post$rkm
post_site <- meta_post$Sampling_Site

onmy_genind_post_location <- df2genind(geno_n_post,
                         sep="/",
                         ind.names=post_ind,
                         loc.names=post_loci, 
                         pop = post_location,
                         ploidy = 2)

onmy_genind_post_site <- df2genind(geno_n_post,
                         sep="/",
                         ind.names=post_ind,
                         loc.names=post_loci, 
                         pop = post_site,
                         ploidy = 2)
onmy_genind_post_location
onmy_genind_post_site
```


# Part 2: Hardy-Weinberg equilibrium  - AS OF 17 May 2022 HWE was checkin in "steeloutliertest.Rmd"


# Part 4: convet genind to genlight objects 

```{r, message=FALSE}
#install_github("green-striped-gecko/dartR")
library(dartR)

genlit_pre_site <- gi2gl(onmy_genind_pre_site)
genlit_pre_location <- gi2gl(onmy_genind_pre_location)
genlit_post_site <- gi2gl(onmy_genind_post_site)
genlit_post_location <- gi2gl(onmy_genind_post_location)
```

### consider filtering genlight objects prior to running a PCA or PCoA?  
prep for pca 
(a) Filter stringently on call rate, using a threshold of at least 95% loci called.
(b) Remove individuals for which call rate is exceptionally low, say <80%.
(c) Impute the remaining missing values on a population‐by‐population basis, where populations can be considered panmictic.
```{r}
#genlit_filter <- gl.filter.callrate(genlit_pre_site, method="loc", threshold=0.90)
#genlit_filter <- gl.filter.callrate(genlit_filter, method="ind", threshold=0.80)
#genlit_filter <- gl.impute(genlit_filter, method="random")
#pcoa <- gl.pcoa(genlit_filter)
```


## PCA using genlight object
set color palette
```{r}
library(viridisLite)
mycol = viridisLite::viridis(4, alpha = 0.6)
```

```{r}
pca_pre <- glPca(genlit_pre_location, center = T, scale = F, nf = 50)
s.class(pca_pre$scores, pop(genlit_pre_location),
        xax=1, yax=2, col=mycol,
        axesel=FALSE, cstar=0, cpoint=3)
```



```{r}
pca_post <- glPca(genlit_post_location, center = T, scale = F, nf = 50)
s.class(pca_post$scores, pop(genlit_post_location),
        xax=1, yax=2, col=mycol,
        axesel=FALSE, cstar=0, cpoint=3)
```


# Part 5: DAPC

i am following this tutorial: https://adegenet.r-forge.r-project.org/files/tutorial-dapc.pdf

find clusters
```{r}
grp_pre <- find.clusters(genlit_pre_location, max.n.clust=40, n.pca=100)
```

check Kstat
```{r}
grp_pre$Kstat
```

will use K=2 and K=3 since BIC stops decreasing all that much beyond there... 
```{r}
grp_pre_2 <- find.clusters(genlit_pre_location, max.n.clust=40, n.clust = 2)
grp_pre_3 <- find.clusters(genlit_pre_location, max.n.clust=40, n.clust = 3)
```

run dapc
```{r}
dapc_pre_2 <- dapc(genlit_pre_location, grp_pre_2$grp, n.pca = 40, n.da = 10) #keep first 40 PCs; and 10 n.da
dapc_pre_3 <- dapc(genlit_pre_location, grp_pre_3$grp, n.pca = 40, n.da = 10) #keep first 40 PCs; and 10 n.da
```

dapc output
```{r}
dapc_pre_2
dapc_pre_3
```

plot DAPC using scatter()
```{r}
scatter(dapc_pre_2, col = viridis(2, alpha = 0.6))
scatter(dapc_pre_3, col = viridis(3, alpha = 0.6))
```

plot DAPC using ggplot
```{r}
dapc_pre_2_scores <- as.data.frame(dapc_pre_2$tab)
dapc_pre_2_scores$Assigned_Pop <- dapc_pre_2$grp
dapc_pre_2_scores$Original_Pop <- meta_pre$Location

set.seed(9)
dapc_p2 <- ggplot(dapc_pre_2_scores, aes(x=PC1, y=PC2, colour=Assigned_Pop)) 
dapc_p2 <- dapc_p2 + geom_point(size=2, aes(shape = Original_Pop)) 
dapc_p2 <- dapc_p2 + stat_ellipse(level = 0.95, size = 1)
dapc_p2 <- dapc_p2 + scale_color_manual(values = viridis(3)) 
dapc_p2 <- dapc_p2 + geom_hline(yintercept = 0) 
dapc_p2 <- dapc_p2 + geom_vline(xintercept = 0) 
dapc_p2 <- dapc_p2 + theme_bw()

dapc_p2
```

plot DAPC using ggplot
```{r}
dapc_pre_3_scores <- as.data.frame(dapc_pre_3$tab)
dapc_pre_3_scores$Assigned_Pop <- dapc_pre_3$grp
dapc_pre_3_scores$Original_Pop <- meta_pre$Location

set.seed(9)
dapc_p3 <- ggplot(dapc_pre_3_scores, aes(x=PC1, y=PC2, colour=Assigned_Pop)) 
dapc_p3 <- dapc_p3 + geom_point(size=2, aes(shape = Original_Pop)) 
dapc_p3 <- dapc_p3 + stat_ellipse(level = 0.95, size = 1)
dapc_p3 <- dapc_p3 + scale_color_manual(values = viridis(3)) 
dapc_p3 <- dapc_p3 + geom_hline(yintercept = 0) 
dapc_p3 <- dapc_p3 + geom_vline(xintercept = 0) 
dapc_p3 <- dapc_p3 + theme_bw()
dapc_p3
```





which loci have the greatest contribution? 
```{r}
set.seed(4)
loadingplot(dapc_pre_2$var.contr, axis=1,
                       thres=.07, lab.jitter=1)
loadingplot(dapc_pre_3$var.contr, axis=1,
                       thres=.07, lab.jitter=1)
loadingplot(dapc_pre_3$var.contr, axis=2,
                       thres=.07, lab.jitter=1)
```


look at loci with >0.05 contribution to pc axis

```{r}
data.frame(dapc_pre_3$var.contr) %>%
  filter(LD2 > 0.05)
```
```{r}
pre_loci[146]
```

## this is Omy_RAD3209-10 on omy 18 (Adaptive. Basin-wide, top-outlier)



```{r}
#summary(dapc_pre_2)
#summary(dapc_pre_3)
```


genotype compostion plot  - individuals are ordered from BD to AD 
```{r}
compoplot(dapc_pre_2, posi="bottomright",
          txt.leg=paste("Cluster", 1:2), xlab="individuals", col=viridis(2))
compoplot(dapc_pre_3, posi="bottomright",
          txt.leg=paste("Cluster", 1:3), xlab="individuals", col=viridis(3))
```


make easier to interpret DAPC plot using ggplot
```{r}
dapc.results <- as.data.frame(dapc_pre_2$posterior)
dapc.results$pop <- meta_pre$Location
dapc.results$indNames <- rownames(dapc.results)
dapc.results <- pivot_longer(dapc.results, -c(pop, indNames))  #pivot table to be able to use ggplot 
```

rename columns and plot
```{r}
colnames(dapc.results) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

p2 <- ggplot(dapc.results, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
p2 <- p2 + geom_bar(stat='identity') 
p2 <- p2 + scale_fill_manual(values = viridis(3)) 
p2 <- p2 + facet_grid(~Original_Pop, scales = "free")
p2 <- p2 + theme(axis.text.x = element_blank())
p2
```

make easier to interpret DAPC plot using ggplot
```{r}
dapc.results <- as.data.frame(dapc_pre_3$posterior)
dapc.results$pop <- meta_pre$Location
dapc.results$indNames <- rownames(dapc.results)
dapc.results <- pivot_longer(dapc.results, -c(pop, indNames))  #pivot table to be able to use ggplot 
```


rename columns and plot
```{r}
colnames(dapc.results) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

p3 <- ggplot(dapc.results, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
p3 <- p3 + geom_bar(stat='identity') 
p3 <- p3 + scale_fill_manual(values = viridis(3)) 
p3 <- p3 + facet_grid(~Original_Pop, scales = "free")
p3 <- p3 + theme(axis.text.x = element_blank())
p3
```


make combined figure for predam individuals... 
```{r}
library(cowplot)

pre_plot <- plot_grid(dapc_p2, p2, dapc_p3, p3, labels = "AUTO", ncol =2)
pre_plot

#ggsave2("outputs/pre_plot.png", width = 14, height = 8)
```




# Now analyses post-removal dam individuals 

find clusters
```{r}
grp_post <- find.clusters(genlit_post_location, max.n.clust=40, n.pca=100)
```

check Kstat
```{r}
grp_post$Kstat
```


will use K=2 to K=5 since BIC stops decreasing all that much beyond there... 
```{r}
grp_post_2 <- find.clusters(genlit_post_location, max.n.clust=40, n.clust = 2)
grp_post_3 <- find.clusters(genlit_post_location, max.n.clust=40, n.clust = 3)
grp_post_4 <- find.clusters(genlit_post_location, max.n.clust=40, n.clust = 4)
grp_post_5 <- find.clusters(genlit_post_location, max.n.clust=40, n.clust = 5)
```

run dapc
```{r}
dapc_post_2 <- dapc(genlit_post_location, grp_post_2$grp, n.pca = 40, n.da = 10) #keep first 40 PCs; and 10 n.da
dapc_post_3 <- dapc(genlit_post_location, grp_post_3$grp, n.pca = 40, n.da = 10) #keep first 40 PCs; and 10 n.da
dapc_post_4 <- dapc(genlit_post_location, grp_post_4$grp, n.pca = 40, n.da = 10) #keep first 40 PCs; and 10 n.da
dapc_post_5 <- dapc(genlit_post_location, grp_post_5$grp, n.pca = 40, n.da = 10) #keep first 40 PCs; and 10 n.da
```


dapc outputs
```{r}
dapc_post_2
dapc_post_3
dapc_post_4
dapc_post_5
```


plot DAPC using scatter()
```{r}
scatter(dapc_post_2, col = viridis(2, alpha = 0.6))
scatter(dapc_post_3, col = viridis(3, alpha = 0.6))
scatter(dapc_post_4, col = viridis(4, alpha = 0.6))
scatter(dapc_post_5, col = viridis(5, alpha = 0.6))
```

plot DAPC using ggplot
```{r}
dapc_post_3_scores <- as.data.frame(dapc_post_3$tab)
dapc_post_3_scores$Assigned_Pop <- dapc_post_3$grp
dapc_post_3_scores$Original_Pop <- meta_post$Location

set.seed(9)
dapc_d3 <- ggplot(dapc_post_3_scores, aes(x=PC1, y=PC2, colour=Assigned_Pop)) 
dapc_d3 <- dapc_d3 + geom_point(size=2, aes(shape = Original_Pop)) 
dapc_d3 <- dapc_d3 + stat_ellipse(level = 0.95, size = 1)
dapc_d3 <- dapc_d3 + scale_color_manual(values = viridis(4)) 
dapc_d3 <- dapc_d3 + geom_hline(yintercept = 0) 
dapc_d3 <- dapc_d3 + geom_vline(xintercept = 0) 
dapc_d3 <- dapc_d3 + theme_bw()
dapc_d3
```

plot DAPC using ggplot
```{r}
dapc_post_4_scores <- as.data.frame(dapc_post_4$tab)
dapc_post_4_scores$Assigned_Pop <- dapc_post_4$grp
dapc_post_4_scores$Original_Pop <- meta_post$Location

set.seed(9)
dapc_d4 <- ggplot(dapc_post_4_scores, aes(x=PC1, y=PC2, colour=Assigned_Pop)) 
dapc_d4 <- dapc_d4 + geom_point(size=2, aes(shape = Original_Pop)) 
dapc_d4 <- dapc_d4 + stat_ellipse(level = 0.95, size = 1)
dapc_d4 <- dapc_d4 + scale_color_manual(values = viridis(4)) 
dapc_d4 <- dapc_d4 + geom_hline(yintercept = 0) 
dapc_d4 <- dapc_d4 + geom_vline(xintercept = 0) 
dapc_d4 <- dapc_d4 + theme_bw()
dapc_d4
```



genotype compostion plot for 2 to 5 clusters 
```{r}
compoplot(dapc_post_2, posi="bottomright",
          txt.leg=paste("Cluster", 1:2), xlab="individuals", col=viridis(2))

compoplot(dapc_post_3, posi="bottomright",
          txt.leg=paste("Cluster", 1:3), xlab="individuals", col=viridis(3))

compoplot(dapc_post_4, posi="bottomright",
          txt.leg=paste("Cluster", 1:4), xlab="individuals", col=viridis(4))

compoplot(dapc_post_5, posi="bottomright",
          txt.leg=paste("Cluster", 1:5), xlab="individuals", col=viridis(5))
```

make easier to interpret membership probability plot using ggplot
```{r}
dapc.results.d3 <- as.data.frame(dapc_post_3$posterior)
dapc.results.d3$pop <- meta_post$Location
dapc.results.d3$indNames <- rownames(dapc.results.d3)
dapc.results.d3  <- pivot_longer(dapc.results.d3, -c(pop, indNames))  #pivot table to be able to use ggplot 
```


rename columns and plot
```{r}
colnames(dapc.results.d3) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

d3 <- ggplot(dapc.results.d3, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
d3 <- d3 + geom_bar(stat='identity') 
d3 <- d3 + scale_fill_manual(values = viridis(4)) 
d3 <- d3 + facet_grid(~Original_Pop, scales = "free")
d3 <- d3 + theme(axis.text.x = element_blank())
d3
```



make easier to interpret membership probability plot using ggplot
```{r}
dapc.results.d4 <- as.data.frame(dapc_post_4$posterior)
dapc.results.d4$pop <- meta_post$Location
dapc.results.d4$indNames <- rownames(dapc.results.d4)
dapc.results.d4  <- pivot_longer(dapc.results.d4, -c(pop, indNames))  #pivot table to be able to use ggplot 
```


rename columns and plot
```{r}
colnames(dapc.results.d4) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

d4 <- ggplot(dapc.results.d4, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
d4 <- d4 + geom_bar(stat='identity') 
d4 <- d4 + scale_fill_manual(values = viridis(4)) 
d4 <- d4 + facet_grid(~Original_Pop, scales = "free")
d4 <- d4 + theme(axis.text.x = element_blank())
d4
```

make combined figure for post dam samples ... 
```{r}
library(cowplot)

post_plot <- plot_grid(dapc_d3, d3, dapc_d4, d4, labels = "AUTO", ncol =2)
post_plot

#ggsave2("outputs/post_plot.png", width = 14, height = 8)
```

